Data Specialization for a Staged Lambda Calculus

### 2015-02-26 paper/splitting.tex
A bunch of minor editing operations in 4.0

In this section, we present a \emph{splitting algorithm} which
statically converts a term in \lang\ into a pair of separate,
monostaged terms that individually correspond to the first and second
stage computations of the \lang\ term. Due to important differences in
structure, we describe both the splitting correctness condition and
splitting algorithm for \bbonem\ terms separately from that of
\bbtwo\ terms.  By construction, \bbonep\ terms are monostage and do
not require splitting.

\textbf{Correctness condition for \bbtwo\ term splitting.}  As
suggested informally in \ref{sec:example}, splitting a \lang\ program
(a \bbtwo\ term) should yield two monostage terms where evaluating the
second term (corresponding to second stage computations) on the output
of the first term (first stage computations), yields the same result
as direct evaluation of the original \lang\ program.  More precisely,
for any $\coltwo e A$ which reduces to a residual $q$ (via $\diatwo e
q$), $e$ splits (via the new relation~$\splittwosym$) into a
syntactically separated program $\pipeS p l r$ where $\reduce p b$,
and $[b/l]r$ is the same as $q$ up to an equivalence. The term $p$
generated by the split (called the {\em precomputation}, due to its
role producing the inputs for subsequent second stage computation) is
a monostage term representing all the first-stage computations in
$e$. The term $l.r$ (called the {\em resumer}) is a monostage term
representing all the second-stage computations in $e$.  We refer to
the result of the precomputation ($b$) the {\em boundary value} of the
split since it represents the data communicated between the first and
second stage computations.  The mechanics of splitting \bbtwo\ terms
and their relationship to the outputs of direct \lang\ evaluation is
provided at right in \ref{fig:splittingSummary}.

@@

To resolve the mismatch between the multi-stage computations
represented by residual tables and partial values and the desired
monostaged outputs of splitting, we choose to define a translation,
called {\em masking}, written $\rtab \xi v \vsplito \mval i q$, which
converts residual tables and partial values into a syntactically
separated {\em masked value}.  The first part of the masked value,
$i$, is a value holding the result of all first stage components of
$v$.  The second component, $q$, is a residual representing all second
stage computations in $\xi$ and $v$.

We define the requirements for splitting \bbonem\ terms using the
results of masking. Specifically, for any $\colmix e A$ which reduces
to $\rtab \xi v$ (via $\redonesym$), masking yields the masked value
$\mval i q$ (via~$\vsplito$).  Then $e$ splits (via the new
relation~$\splitonesym$) into a syntactically separated program
$\pipeM c l r$ where $\reduce c (i,b)$, and $[b/l]r$ is the same as
$q$ up to an equivalence.  The term $c$ generated by splitting (called
the {\em combined term}) is a monostage term encoding all the
first-stage computations in $e$, and $l.r$ (again called the {\em
  resumer}) is a monostage term encoding all the second-stage
computations in $e$.  A summary of splitting \bbonem\ terms is
provided at left in \ref{fig:splittingSummary}.

@@

Note that since $\coltwo e A$ reduces to an entirely second-stage
residual, its first-stage computations only exist to generate input
for second stage computations.  In contrast, since $\colmix e A$
reduces to a multi-stage residual table and partial value, its first
stage computations serve to both generate the value used as input for
the second stage {\em and also} produce a first-stage value.
Correspondingly, the precomputation $p$ generated when splitting
\bbtwo\ terms reduces only to a boundary value $b$ which is passed to
the resumer; but the combined term $c$ resulting from \bbonem\ term
splitting reduces to a tuple containing both the first stage result
$i$ and the boundary value $b$ (motivating the name {\em combined
term}).

@@

\subsection{Masking}

>>>>

In this section, we present a {\em splitting algorithm} that
statically converts a term in \lang\ into a pair of monostaged terms
that individually correspond to the first and second stage
computations of the original \lang\ term. Due to important differences
in structure, we describe both the splitting correctness condition and
splitting algorithm for \bbonem\ terms separately from that of
\bbtwo\ terms.  By construction, \bbonep\ terms are monostage and do
not require splitting.

\textbf{Correctness condition for \bbtwo\ term splitting.}  As
suggested informally in \ref{sec:example}, splitting a \lang\ program
(a \bbtwo\ term) should yield two monostage terms where evaluating the
second term (corresponding to second stage computations) on the output
of the first term (first stage computations), yields the same result
as direct evaluation of the original \lang\ program.  More precisely,
for any $\coltwo e A$ which reduces to a residual $q$ (via $\diatwo e
q$), $e$ splits (via the new relation~$\splittwosym$) into the
syntactically separated terms $\pipeS p l r$ where $\reduce p b$, and
$[b/l]r$ is the same as $q$ up to an equivalence. The term $p$
generated by the split (called the {\em precomputation}, due to its
role producing the inputs for subsequent second stage computation) is
a monostage term representing all the first-stage computations in
$e$. The term $l.r$ (called the {\em resumer}) is a monostage term
representing all the second-stage computations in $e$.  We refer to
the result of the precomputation ($b$) the {\em boundary value} of the
split since it represents the data communicated between the first and
second stage computations.  The process of splitting \bbtwo\ terms
(and the relationship of these operations to the outputs of direct
\lang\ evaluation) is provided at right in \ref{fig:splittingSummary}.
% and described in detail in \ref{sec:split-two}.

@@

To resolve the mismatch between the multi-stage computations
represented by residual tables and partial values and the desired
monostaged outputs of splitting, we choose to define a translation,
called {\em masking}, written $\rtab \xi v \vsplito \mval i q$, which
converts residual tables and partial values into a syntactically
separated {\em masked value}.  The first part of the masked value,
$i$, is a monostage value holding the result of all first stage
components of $v$.  The second component, $q$, is a residual
representing all second stage computations in $\xi$ and $v$. (A precise definition of masking is given in \ref{sec:masking}.)

We define the requirements for splitting \bbonem\ terms using the
results of masking. Specifically, for any $\colmix e A$ which reduces
to $\rtab \xi v$ (via $\redonesym$), masking yields the masked value
$\mval i q$ (via~$\vsplito$).  Then $e$ splits (via the new
relation~$\splitonesym$) into the monostage terms $\pipeM c l r$ where
$\reduce c (i,b)$, and $[b/l]r$ is the same as $q$ up to an
equivalence.  The term $c$ generated by splitting (called the {\em
  combined term}) is a monostage term encoding all the first-stage
computations in $e$, and $l.r$ (again called the {\em resumer}) is a
monostage term encoding all the second-stage computations in $e$.  The
process of splitting \bbonem\ terms is summarized at left in
\ref{fig:splittingSummary}.
% and described in detail in \ref{sec:split-one}.

@@

Note that since $\coltwo e A$ reduces to an entirely second-stage
residual, its first-stage computations only exist to generate input
for second stage computations.  In contrast, since $\colmix e A$
reduces to a multi-stage residual table and partial value, its first
stage computations serve to both generate the value used as input for
the second stage {\em and also} produce a first-stage value.
Correspondingly, the precomputation $p$ resulting from splitting
\bbtwo\ terms reduces only to a boundary value $b$ which is passed to
the resumer; but the combined term $c$ resulting from splitting
\bbonem\ terms reduces to a tuple containing both the first stage {\em
  immediate result} $i$ and the boundary value $b$ (motivating the
name {\em combined terxm}).

@@

\subsection{Masking}
\label{sec:masking}
